---
title: "607 Final Project - Cause of Death by Firearm vs State Firearm Laws"
author: "Jen Abinette"
date: "2022-11-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Research question 

Are there aspects of gun policy that are predictors of firearm mortality rate?

### Cases 

All datasets and relevant information can be found in Github Repository - https://github.com/JAbinette/CUNY-607-Final-Project

Cause of Death by Firearm data retrieved from Centers for Disease Control & Prevention website includes 458 observations grouped by State, Year and Cause of Death Category (ICD Sub-Chapter) including firearm deaths excluding Terrorism, Legal Intervention and Operations of War 
  See Github File: '0 - wonder.cdc.gov Underlying Cause of Death cdc - Grp by State, ICD Sub-Chapter.txt'

Firearm Laws by State data retrieved from https://www.statefirearmlaws.org/resources Database containing detailed annual information on firearm-related laws in place in each of the 50 US states includes 150 observations with 137 variables after data subset to only include 2018-2020.
  See Github File: '0 - statefirearmlaws.org Firearm Laws - DATABASE_0.xlsx'

### Data Preparation

### CDC 2020 Summary of Death by Firearm

```{r}
library(tidyr)
library(readxl)
library(dplyr)
# Load CDC Cause of Death by Firearm text file from Github
txt.url = 'https://raw.githubusercontent.com/JAbinette/CUNY-607-Final-Project/main/0%20-%20wonder.cdc.gov%20Underlying%20Cause%20of%20Death%20cdc%20-%20Grp%20by%20State%2C%20ICD%20Sub-Chapter.txt'
cod.ld <- read.delim (txt.url, header=TRUE, sep = "\t")

cod.st_yr <- subset (cod.ld, Year.Code == 2020, select = c("State", "Year.Code", "Deaths", "Population")) %>%
  group_by(State, Year.Code) %>%
  mutate(Firearm_Deaths = sum(Deaths)) %>%
  mutate (Prop.of.Pop = (Firearm_Deaths/Population) ) %>%
  ungroup()
cod.st_yr <- subset (cod.st_yr, State != "District of Columbia", select =c("State", "Year.Code", "Firearm_Deaths", "Prop.of.Pop")) %>%
  distinct() %>%
  na.omit()

``` 

### CDC Reported Deaths by State in 2020

```{r}
# Load CDC Deaths by State in 2020 text file from Github
txt.url = 'https://raw.githubusercontent.com/JAbinette/CUNY-607-Final-Project/main/0%20-%20wonder.cdc.gov%202020%20Deaths%20by%20State.txt'
cod.all <- read.delim (txt.url, header=TRUE, sep = "\t")
cod.all <- subset(cod.all, Year.Code == 2020 & State != "District of Columbia", select = c("State", "Deaths"))

# Merge with Firearm Deaths data
COD <- merge( cod.st_yr, cod.all, by.x = "State", by.y = "State", all.x = TRUE)  %>%
  mutate (Prop.of.Deaths = (Firearm_Deaths/Deaths) )

```

### State Firearm Laws (filtered to only include 2020)

```{r}
# Download State Firearm Laws spreadsheet and save to your file directory (data saved in Github at https://github.com/JAbinette/CUNY-607-Final-Project/blob/main/0%20-%20statefirearmlaws.org%20Firearm%20Laws%20-%20DATABASE_0.xlsx)
# Set path to excel spreadsheet
path = "0 - statefirearmlaws.org Firearm Laws - DATABASE_0.xlsx"
laws_ld <- read_excel(path)
# Subset data to only include years 2018-2020
laws_ld <- subset (laws_ld, year == 2020)
# Transform from wide to long (excluding last column which is the total number of laws)
laws_ld2 <- pivot_longer ( laws_ld, cols = 3:136, names_to ="Variable", values_to = "Indicator" )
# Remove variable records where the law was not present for that year and State
laws_ld3 <- subset ( laws_ld2, Indicator == 1 )
head(laws_ld3)
```

### Add Variable Categories to Merge with State Firearm 

```{r}
# Load Variable Category data for State Firearm Laws data 
laws.cat_ld <- read.csv('https://raw.githubusercontent.com/JAbinette/CUNY-607-Final-Project/main/0%20-%20statefirearmlaws.org%20Variable%20Categories%20-%20codebook_0.csv')
# Replace spaces in Category with a period
laws.cat_ld$Category <- make.names (laws.cat_ld$Category)
# Merge
laws_ld4 <- merge( laws_ld3, laws.cat_ld, by.x = "Variable", by.y = "Variable.Name", all.x = TRUE)
laws_ld4 <- subset ( laws_ld4, select = c("state","year","lawtotal","Category"))

library(dplyr)
# Add count of laws by Category 
laws_ld4 <- laws_ld4 %>%
  group_by(state, year, lawtotal, Category) %>%
  summarise(Category_count=n(),
            .groups = 'drop')
head(laws_ld4)
```

```{r}
# Transform long to wide
laws <- pivot_wider( laws_ld4, names_from = "Category", values_from = "Category_count")
# Convert NA to zero
laws[is.na(laws)] <- 0

# Merge State Laws data with Cause of Death by State and Year
df <- merge (COD, laws, by.x = c("State"), by.y = c("state"), all.x = TRUE)
df <- subset(df, select = c("State", "Prop.of.Pop", "Prop.of.Deaths", "lawtotal"))
```

# Statistical Analyses 

```{r}
summary(df)
```

### Can the Total Number of Laws predict proportion of Firearm deaths per State Population?

```{r}
df %>%
  summarise( cor( lawtotal, Prop.of.Pop, use = "complete.obs"))
library(ggplot2)
ggplot(data = df, aes(x = lawtotal, y = Prop.of.Pop)) + geom_jitter() + stat_smooth(method = "lm", se = FALSE)
```


```{r}
# Simple Linear Regression
m.pop <- lm(Prop.of.Pop ~ lawtotal, data = df)
summary(m.pop)
```

### Check Model Assumptions for Simple Regression

### Linearity and Constant Variability - Conditions both met as there is no apparent pattern in the residuals plot indicating there is linearity and the points are scattered around zero showing constant variability.
```{r}
ggplot(data = m.pop, aes(x = .fitted, y = .resid)) +
geom_point() +
geom_hline(yintercept = 0, linetype = "dashed") +
xlab("Fitted values") +
ylab("Residuals")
```

### Nearly Normal Residuals - Condition is met based on below histogram and the normal probability plots.
```{r}
ggplot(data = m.pop, aes(x = .resid)) +
geom_histogram(binwidth = .00001) +
xlab("Residuals")

ggplot(data = m.pop, aes(sample = .resid)) + stat_qq()
```

### Can the Number and Category of Firearm Laws predict proportion of Firearm deaths per State Population?
```{r}
# Merge State Laws data with Cause of Death by State and Year
df2 <- merge (COD, laws, by.x = c("State"), by.y = c("state"), all.x = TRUE)

# Multiple Linear Regression
m_2020 <- lm(Prop.of.Pop ~ Concealed.carry.permitting + Dealer.regulations + Domestic.violence + 
Prohibitions.for.high.risk.gun.possession + Buyer.regulations + Possession.regulations + Stand.your.ground + Ammunition.regulations + Assault.weapons.and.large.capacity.magazines + 
Background.checks + Child.access.prevention + Gun.trafficking + Immunity + Preemption, data = df2)
summary(m_2020)
```

### Fit the Best Model using backward-selection and p-value as the selection criterion
```{r}
# Variables removed highest p-value at a time: Immunity, Dealer.regulations, Concealed.carry.permitting, Background.checks, Prohibitions.for.high.risk.gun.possession, Domestic.violence, Assault.weapons.and.large.capacity.magazines, Gun.trafficking,  Ammunition.regulations, Buyer.regulations, Possession.regulations
m_bestfit <- lm(Prop.of.Pop ~ Stand.your.ground + Child.access.prevention + Preemption, data = df2)
summary(m_bestfit)
```

### Check Model Assumptions

```{r}
par(mfrow = c(2, 2))
plot(m_bestfit)
```

### Conclusions

The total number of firearm laws is a significant predictor and accounts for 49% of the variability in Firearm Deaths as a Proportion of the State Population. In breaking down the firearm laws by category, a multiple regression analysis determined that the number of Stand Your Ground, Child Access Prevention, and Preemption firearm laws can explain 55% of the variability.

